Live Functional Programming with Typed Holes

### 2018-04-14 calculus.tex
another page down

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Matt is here (2018.04.14, ~6:15am MT)

@@

%%%%%%%%%%%%%%%%%%% MATT is here (2018.04.14 -- 6:40am MT)

The dynamic semantics is specified as a ``small-step'' transition system. 
The cast-related machinery is based closely on the cast calculus from the ``refined'' account of the gradually typed lambda calculus by \citet{DBLP:conf/snapl/SiekVCB15}, which is known to be theoretically well-behaved. 
In particular, Fig.~\ref{fig:isGround} defines the judgement $\isGround{\htau}$, which distinguishes the base type, $b$, and the least specific arrow type, $\tarr{\tehole}{\tehole}$, as \emph{ground types}, which play a role in simplifying the treatment of function casts. 
Fig.~\ref{fig:isFinal} defines the judgement $\isFinal{d}$, which distinguishes the final forms of the transition system. 
There are two classes of final forms: (possibly-)boxed values and indeterminate forms.%
\footnote{In most accounts of the cast calculus, values and ground types are distinguished with separate grammars together with an implicit identification convention. 
Our judgemental formulation is more faithful to the mechanization and cleaner for our purposes, because we are distinguishing several classes of final forms.}

@@

The judgement $\isBoxedValue{d}$ distinguishes (possibly-)boxed values, which correspond to the values from the cast calculus and include the classic values from the lambda calculus, distinguished by $\isValue{d}$, as well as two cast forms: 
casts between function types (except identity casts) 
and casts from a ground type to the hole type. In both cases, the cast must be on a boxed value.

The judgement $\isIndet{d}$ distinguishes \emph{indeterminate} forms, 
so named because they arise from the presence of expression holes and failed casts in a program, and so, conceptually, their ultimate value awaits programmer action (see Sec.~\ref{sec:resumption}). 
The first two rules specify that hole closures are indeterminate (in the case of non-empty hole closures, when the inner expression is final). 
The rules defining the remaining indeterminate forms are explained simultaneously with the corresponding transition rules below.\todo{relationship to weak head normal forms}

@@

The transition rules are defined in Fig.~\ref{fig:instruction-transitions}-\ref{fig:step}. 
Top-level transitions, i.e. \emph{steps}, $\stepsToD{}{d}{d'}$, are governed by a single rule, Rule {Step} in Fig.~\ref{fig:step}, which 
(1) decomposes $d$ into an evaluation context, $\evalctx$, and a selected sub-term, $d_0$; 
(2) takes an \emph{instruction transition}, $\reducesE{}{d_0}{d_0'}$, as specified in Fig.~\ref{fig:instruction-transitions}; 
and (3) places $d_0'$ back in the selected position, which is marked%
\footnote{In the literature, the form $\evalhole$ in the grammar of evaluation contexts is referred to as the \emph{hole}, but this hole is a technical device entirely orthogonal to the holes of this paper, so we use the term ``mark'' instead.} 
in the evaluation context by $\evalhole$, to obtain $d'$.
This approach was originally developed in the reduction semantics of \citet{DBLP:journals/tcs/FelleisenH92} and is the predominant style of operational semantics in the literature on gradual typing. 
Because we distinguish final forms judgementally, rather than syntactically, we use a judgemental formulation of this approach called a \emph{contextual dynamics} by \citet{pfpl}. 
It would be straightforward to derive an equivalent structural operational semantics \cite{DBLP:journals/jlp/Plotkin04a} by using search rules instead of evaluation contexts (\citet{pfpl} relates the two approaches).
\todo{Put the search rules in the appendix?}{}

>>>>

@@

The dynamic semantics consists of a ``small-step'' transition system,
%
where cast-related machinery closely follows the cast calculus from
the ``refined'' account of the gradually typed lambda calculus
by \citet{DBLP:conf/snapl/SiekVCB15}, which is known to be
theoretically well-behaved.
%
In particular, \Figref{fig:isGround} defines the judgement
$\isGround{\htau}$, which distinguishes the base type~$b$ and the
least specific arrow type~$\tarr{\tehole}{\tehole}$, as \emph{ground
types}; this judgement helps simplify the treatment of function casts.
%
\Figref{fig:isFinal} defines the judgement $\isFinal{d}$, which
distinguishes the final forms of the transition system.
%
There are two classes of final forms: (possibly-)boxed values and
indeterminate forms.%
%
\footnote{
        In most accounts of the cast calculus, values and ground types
        are distinguished with separate grammars together with an
        implicit identification convention.
        %
        Our judgemental formulation is more faithful to the mechanization and
        cleaner for our purposes, because we are distinguishing several
        classes of final forms.
}

@@

The judgement $\isBoxedValue{d}$ defines (possibly-)boxed values as
either ordinary values~(Rule~{BVVal}), or one of two cast forms: casts
between function types (except identity casts) and casts from a ground
type to the hole type. In each case, the cast operates on a boxed
value.
%
%, which correspond to the values from the cast calculus and include
%the classic values from the lambda calculus, distingui%shed by
%$\isValue{d}$,
%
The judgement $\isIndet{d}$ defines \emph{indeterminate} forms, so
named because they arise from the presence of expression holes and
failed casts in a program, and so, conceptually, their ultimate value
awaits programmer action (see Sec.~\ref{sec:resumption}).
%
The first two rules specify that \emph{empty} hole closures are always
indeterminate, and that \emph{non}-empty hole closures are indeterminate when
they consist of a \emph{final} inner expression.
%
Below, we discuss the the remaining indeterminate forms simultaneously
with the corresponding transition rules, also discussed
below.\todo{relationship to weak head normal forms}

@@

The transition rules are defined in Fig.~\ref{fig:instruction-transitions}-\ref{fig:step}.
%
Top-level transitions are \emph{steps}~$\stepsToD{}{d}{d'}$, governed by Rule~{Step} in \Figref{fig:step}, which
%
(1) decomposes $d$ into an evaluation context, $\evalctx$, and a selected sub-term, $d_0$;
%
(2) takes an \emph{instruction transition}, $\reducesE{}{d_0}{d_0'}$, as specified in \Figref{fig:instruction-transitions};
%
and (3) places $d_0'$ back in the selected position, indicated in
the evaluation context by the \emph{mark}~$\evalhole$, to obtain $d'$.%
\footnote{
        We say ``mark'' here to avoid confusion with the (orthogonal) ``holes'' of \HazelnutLive.
        %
        %the form $\evalhole$ in the grammar of
        %evaluation contexts is referred to as the \emph{hole}, but
        %this hole is a technical device entirely orthogonal to the
        %holes of this paper, so we use the term ``mark'' instead.
        }
%

\matt{Next paragraph can be dropped entirely for space}
%
This approach was originally developed in the reduction semantics of \citet{DBLP:journals/tcs/FelleisenH92} and is the predominant style of operational semantics in the literature on gradual typing. 
Because we distinguish final forms judgementally, rather than syntactically, we use a judgemental formulation of this approach called a \emph{contextual dynamics} by \citet{pfpl}. 
It would be straightforward to derive an equivalent structural operational semantics \cite{DBLP:journals/jlp/Plotkin04a} by using search rules instead of evaluation contexts (\citet{pfpl} relates the two approaches).
\todo{Put the search rules in the appendix?}{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% MATT is here (2018.04.14 -- ~7am MT)
